Generating Hubbard Model Solutions from Anderson Impurity Model Solutions

### 2021-08-12 notes.tex
changed unitary to similarity, added figure for the improved wavefunction

\title{Generating Hubbard Model Solutions from Anderson Impurity Model Solutions}
\author{}%Abhirup Mukherjee, Dr. Siddhartha Lal}
\begin{document}
\maketitle
\section{Introduction}
This is an attempt to obtain various quantities like Greens functions, self-energies, spectral functions and (if possible) energies and wavefunctions of the Hubbard model, using a cluster-bath approach. The cluster-bath system is taken to be a single-impurity Anderson model with a correlated bath. The correlation will be brought about in two ways: a self-energy $\Sigma(k,\omega)$ of the bath, and a double occupancy repulsion cost $U_b$. The Hubbard and the correlated single-impurity Anderson models are defined using the Hamiltonians
\begin{align}
H_\text{hubb} &= -t^H\sum_{\sigma,\left<i,j \right>}\left(c^\dagger_{i\sigma} c_{j\sigma} + \text{h.c.}\right) + U^H\sum_i \hat n_{i \uparrow} \hat n_{i \downarrow} - \mu^H \sum_{i\sigma}\hat n_{i\sigma}\\
H_\text{siam} &= \sum_{k\sigma}\left[\epsilon_k + \Sigma(k,\omega)\right]\hat n_{k\sigma} + \epsilon_d^A \sum_\sigma\hat n_{d\sigma} + U^A \hat n_{d \uparrow} \hat n_{d \downarrow} + U_b \sum_{kk^\prime}\hat n_k \hat n_{k^\prime} - \mu^A \sum_{i\neq d,\sigma}\hat n_{i\sigma} \label{clus_bath_siam}
\end{align}
Broadly speaking, the method involves first solving the SIAM using a unitary renormalisation group approach, to get the low energy effective theory, and then combining the low energy Hamiltonians in a symmetrized fashion to get the Hamiltonian for the Hubbard model lattice. It is reminescent of dynamical mean-field theory (DMFT) - both involve an impurity-solver that solves an auxiliary system. The difference, however, lies in the following points:
\begin{itemize}
	\item While DMFT primarily works with Greens functions and self-energies, this method involves Hamiltonians. The impurity-solver in DMFT provides an impurity Greens function (which is then equated with the local Greens function of the bath), while the impurity-solver in this method actually provides a low energy Hamiltonian.
	\item The final step of DMFT is the self-consistency equation, where the impurity and bath-local quantities are set equal. This ensures all sites, along with the impurity site, have the same self-energy, something which is required on grounds of  translational invariance. The present method, however, brings about the translational invariance in a different way. It symmetrizes the Hamiltonians itself, such that all quantites then derived from the Hamiltonian are then guaranteed to have the symmetry.
\end{itemize}
The meaning of each of these statements will become clearer when we describe the method in more detail.

@@

	\tilde H = \mathcal{U} H^A \mathcal{U}^\dagger

@@

The content of the above equation is that we expect the final tiled translationally invariant new (renormalised) Hubbard Hamiltonian to be connected to the bare parent Hubbard model through some unitary transformation $\mathcal{U}$.

@@

The claim made in a previous section was that this Hamiltonian is very close to a unitarily transformed Hubbard Hamiltonian:

@@

	\frac{2}{Nw}\sum_{\left<ij \right>} H^D(i,j,t^D, U^D) \simeq  \mathcal{U} H^H  \mathcal{U}^\dagger

@@

$\mathcal{U}$ is actually distinct from the unitary transformation that completes the URG starting from the bare cluster+bath Anderson model up to the fixed-point Hamiltonian. It is a larger transformation that involves three steps: the URG transformations $\mathcal{U}_A$, the extracting of the zero mode $Z$ and the tiling procedure outline in this section, $T$.

@@

	\mathcal{U} = T Z \mathcal{U}_A

@@

	\left(G_H\right)^\sigma_{ij}(t) \equiv -i\theta(t)\bra{0}\left\{c_{i\sigma},c^\dagger_{j\sigma} \right\}\ket{0} &= -i\theta(t)\bra{0} \mathcal{U}^\dagger\left\{  \mathcal{U} c_{i\sigma}  \mathcal{U}^\dagger  \mathcal{U} c^\dagger_{j\sigma}  \mathcal{U}^\dagger +  \mathcal{U} c^\dagger_{j\sigma}  \mathcal{U}^\dagger  \mathcal{U} c_{i\sigma}  \mathcal{U}^\dagger\right\} \mathcal{U}\ket{0}\\

@@

This says that the Greens function $G_H$ of the bare Hubbard model $H^H$ for excitations $\ket{{j\sigma}} \equiv c^\dagger_{j\sigma}\ket{0}$ on top of its ground state $\ket{0}$ is exactly equal to the Greens function $\tilde G$ of the renormalised Hamiltonian $\tilde H$ for renormalised excitations $\ket{\tilde {j\sigma}} \equiv \tilde c^\dagger_{j\sigma}\ket{\tilde 0} \equiv  \mathcal{U} c^\dagger_i  \mathcal{U}^\dagger  \mathcal{U}\ket{0} =  \mathcal{U}\ket{{j\sigma}}$ on top of its renormalised ground state $\ket{\tilde 0} =  \mathcal{U} \ket{0}$. This means,

@@

	\left[\left(G_H\right)^{-1}\right]^\sigma_{ij} = \bra{i\sigma}\omega - H^H\ket{j\sigma} = \bra{i\sigma} \mathcal{U}^\dagger\left(\omega -  \mathcal{U} H^H  \mathcal{U}^\dagger\right)  \mathcal{U}\ket{j\sigma} = \bra{\tilde i\sigma}\left(\omega - \tilde H\right)\ket{\tilde j\sigma} = \left[\left(\tilde G\right)^{-1}\right]^\sigma_{\tilde i,\tilde j}

@@

\subsection{On the Bethe lattice}
Another test involves considering the case of infinite number of nearest neighbours $w\to\infty$ (the coordination number, and effectively the dimensionality). Here, as has been argued in the DMFT literature, the correct scaling of the $t^{H}$ hopping parameter is $t^{H}\to t^{H}/\sqrt{w}$ such that the kinetic energy of the associated tight-binding lattice model is finite. This allows for the competition between the kinetic and potential terms of the Hamiltonian to drive a metal-insulator transition in the limit of $w\to\infty$ as well. Further, it has been argued in the DMFT literature that the Greens function matrix of the Hubbard model on the Bethe lattice with $w\to\infty$ becomes purely local (i.e., it has vanishing inter-site matrix elements). We can also see this from eqs.\eqref{local_gf},\eqref{nn_gf} where $(G^{H})^{-1}$ matrix has no $k$ dependence as $(G^{H})^{-1}_{ij}\propto 1/w \to 0$ as $w\to\infty$. Thus, in the self-energy 
\begin{eqnarray}
\Sigma_{H} (\vec{k},\omega) &=& \omega - g_{0} + (t^{H} - g_{1})\xi_{\vec{k}}\nonumber\\
&\to & \lim_{w\to\infty} \omega - {g}_{0} + t^{H}\xi_{\vec{k}}~,
\end{eqnarray}
We can now see that even in the limit of $w\to\infty$, the competition between ${g}_{0}$ (the on-site repulsion) and the hopping related kinetic energy ($t^{H}\xi_{\vec{k}}$) can lead to a metal-insulator transition.
\par\noindent
While we have provided expressions for the single particle Greens functions of the $N$-site Hubbard model from their equivalent single particle Greens functions of the Hubbard dimer, we expect that eq.\eqref{green_eq_final} holds quite generally for the two-particle Greens function sector as well. We will provide these expressions at a later point in time. 
\section{Calculating $g_0$ and $g_1$ in practice}
The expressions for $g_0$ and $g_1$, as mentioned above, involve taking matrix elements of the Hubbard dimer Hamiltonian against the exact wavefunctions of the full Hubbard model.
\begin{itemize}
	\item Since the full Hubbard model ground state wavefunctions are not readily available, one can, as the simplest approximation, assume the ground state has the form
	\begin{equation}\begin{aligned}
		\ket{\tilde 0} = \sum_{\left<ij \right>}\ket{\Phi^D_{ij}}\otimes\ket{\Psi_{\hat i, \hat j}}\label{guess}
	\end{aligned}\end{equation}
	$\ket{\Phi^D_{ij}}$ would be the ground state of the Hubbard dimer with $i,j$ creating the two sites of the dimer, while $\ket{\Psi_{\hat i, \hat j}}$ would be the wavefunction involving the rest of the sites. This ofcourse assumes that the two sets $(i,j)$ and $(a,b,...,h,k,...)$ are not entangled, and is not true in general. With this assumption, $g_0$ and $g_1$ become related to the inverse Greens functions of the Hubbard dimer. Such a choice of the wavefunction is motivated by the fact that in the auxiliary system we chose, the bath had only a diagonal interaction; there was no off-diagonal two particle scattering term.
	\item Another way of approaching this is to obtain the ground state wavefunction numerically, and then computing the matrix elements of $H^D$ against this wavefunction to obtain $g_0$ and $g_1$ for each value of $U^H$ and $t^H$ of the bare Hubbard model. This however means we lose any analytical insight into $g_0$ and $g_1$.
	\item The overall best way of calculating $g_0$ and $g_1$ is probably by systematically improving the ground state wavefunctions. This can done by obtaining the wavefunction not just for the dimer, but for a dimer with dispersion:
\begin{equation}\begin{aligned}
		H^D_\text{improved} = -t\sum_{{k\in\text{NN of }z}}\left(c^\dagger_{k\sigma}c_{z\sigma} + \text{h.c.}\right) -t \left(c^\dagger_{i\sigma}c_{z\sigma} + \text{h.c.}\right) + U\left(\tau_{d \uparrow}\tau_{d \downarrow} + \tau_{z \uparrow}\tau_{z \downarrow}\right)
\end{aligned}\end{equation}

@@

By systematically increasing the number of nearest neighbours, we can get better and better ground state wavefunctions, with which we can compute matrix elements, and hence $g_0$, $g_1$. What this means is that if we had just a Hubbard dimer wavefunction, then the impurity can at most hop onto its nearest neighbour, and then hop back, hence the wavefunction in eq.~\ref{guess}. However, with this additional hopping, the impurity can now now hop on to the nearest neighbour $z$, and then on to the next nearest neighbours. This leads to entanglement between the impurity site and the next nearest neighbour sites, which was missing in eq.~\ref{guess}. Another improvement can be made by introducing a self-energy into the dispersion. This introduces correlation on sites beyond the impurity and its nearest neighbour. The introduction of the dispersion and a self-energy gives the hope that we can capture a spectral function of the three peak form, because previously we have seen such a spectral function in the Anderson molecule with dispersion.

>>>>

\title{Generating Hubbard Model Solutions from Anderson Impurity Model Solutions}
\author{}%Abhirup Mukherjee, Dr. Siddhartha Lal}
\begin{document}
\maketitle
\section{Introduction}
This is an attempt to obtain various quantities like Greens functions, self-energies, spectral functions and (if possible) energies and wavefunctions of the Hubbard model, using a cluster-bath approach. The cluster-bath system is taken to be a single-impurity Anderson model with a correlated bath. The correlation will be brought about in two ways: a self-energy $\Sigma(k,\omega)$ of the bath, and a double occupancy repulsion cost $U_b$. The Hubbard and the correlated single-impurity Anderson models are defined using the Hamiltonians
\begin{align}
H_\text{hubb} &= -t^H\sum_{\sigma,\left<i,j \right>}\left(c^\dagger_{i\sigma} c_{j\sigma} + \text{h.c.}\right) + U^H\sum_i \hat n_{i \uparrow} \hat n_{i \downarrow} - \mu^H \sum_{i\sigma}\hat n_{i\sigma}\\
H_\text{siam} &= \sum_{k\sigma}\left[\epsilon_k + \Sigma(k,\omega)\right]\hat n_{k\sigma} + \epsilon_d^A \sum_\sigma\hat n_{d\sigma} + U^A \hat n_{d \uparrow} \hat n_{d \downarrow} + U_b \sum_{kk^\prime}\hat n_k \hat n_{k^\prime} -t^A\sum_{k\sigma}\left(c^\dagger_{d\sigma}c_{k\sigma} + \text{h.c.}\right) \label{clus_bath_siam}
\end{align}
Broadly speaking, the method involves first solving the SIAM using a unitary renormalisation group approach, to get the low energy effective theory, and then combining the low energy Hamiltonians in a symmetrized fashion to get the Hamiltonian for the Hubbard model lattice. It is reminescent of dynamical mean-field theory (DMFT) - both involve an impurity-solver that solves an auxiliary system. The difference, however, lies in the following points:
\begin{itemize}
	\item While DMFT primarily works with Greens functions and self-energies, this method involves Hamiltonians. The impurity-solver in DMFT provides an impurity Greens function (which is then equated with the local Greens function of the bath), while the impurity-solver in this method actually provides a low energy Hamiltonian.
	\item The final step of DMFT is the self-consistency equation, where the impurity and bath-local quantities are set equal. This ensures all sites, along with the impurity site, have the same self-energy, something which is required on grounds of  translational invariance. The present method, however, brings about the translational invariance in a different way. It symmetrizes the Hamiltonians itself, such that all quantites then derived from the Hamiltonian are then guaranteed to have the symmetry.
\end{itemize}
The meaning of each of these statements will become clearer when we describe the method in more detail.

@@

	\tilde H = \mathcal{U} H^H \mathcal{U}^{-1}

@@

where $\mathcal{U}$ is some transformation that is either a unitary, or, at the very least, a similarity transformation.. The content of the above equation is that we expect the final tiled translationally invariant new (renormalised) Hubbard Hamiltonian to be connected to the bare parent Hubbard model through some unitary/similarity transformation $\mathcal{U}$.

@@

The claim made in a previous section was that this Hamiltonian is very close to a similarity transformed Hubbard Hamiltonian:

@@

	\frac{2}{Nw}\sum_{\left<ij \right>} H^D(i,j,t^D, U^D) \simeq  \mathcal{U} H^H  \mathcal{U}^{-1}

@@

$\mathcal{U}$ is in general distinct from the unitary transformation that completes the URG starting from the bare cluster+bath Anderson model up to the fixed-point Hamiltonian. It is a larger transformation that involves three steps: the URG transformations $\mathcal{U}_A$, the extracting of the zero mode $Z$ and the tiling procedure outline in this section, $T$.

@@

	\mathcal{U} = T Z \mathcal{U}

@@

	\left(G_H\right)^\sigma_{ij}(t) \equiv -i\theta(t)\bra{0}\left\{c_{i\sigma},c^\dagger_{j\sigma} \right\}\ket{0} &= -i\theta(t)\bra{0} \mathcal{U}^{-1}\left\{  \mathcal{U} c_{i\sigma}  \mathcal{U}^{-1}  \mathcal{U} c^\dagger_{j\sigma}  \mathcal{U}^{-1} +  \mathcal{U} c^\dagger_{j\sigma}  \mathcal{U}^{-1}  \mathcal{U} c_{i\sigma}  \mathcal{U}^{-1}\right\} \mathcal{U}\ket{0}\\

@@

This says that the Greens function $G_H$ of the bare Hubbard model $H^H$ for excitations $\ket{{j\sigma}} \equiv c^\dagger_{j\sigma}\ket{0}$ on top of its ground state $\ket{0}$ is exactly equal to the Greens function $\tilde G$ of the renormalised Hamiltonian $\tilde H$ for renormalised excitations $\ket{\tilde {j\sigma}} \equiv \tilde c^\dagger_{j\sigma}\ket{\tilde 0} \equiv  \left(\mathcal{U} c^\dagger_i  \mathcal{U}^{-1}\right)  \mathcal{U}\ket{0} =  \mathcal{U}\ket{{j\sigma}}$ on top of its renormalised ground state $\ket{\tilde 0} =  \mathcal{U} \ket{0}$. This means,

@@

	\left[\left(G_H\right)^{-1}\right]^\sigma_{ij} = \bra{i\sigma}\omega - H^H\ket{j\sigma} = \bra{i\sigma} \mathcal{U}^{-1}\left(\omega -  \mathcal{U} H^H  \mathcal{U}^{-1}\right)  \mathcal{U}\ket{j\sigma} = \bra{\tilde i\sigma}\left(\omega - \tilde H\right)\ket{\tilde j\sigma} = \left[\left(\tilde G\right)^{-1}\right]^\sigma_{\tilde i,\tilde j}

@@

\subsection{On the Bethe lattice}
Another test involves considering the case of infinite number of nearest neighbours $w\to\infty$ (the coordination number, and effectively the dimensionality). Here, as has been argued in the DMFT literature, the correct scaling of the $t^{H}$ hopping parameter is $t^{H}\to t^{H}/\sqrt{w}$ such that the kinetic energy of the associated tight-binding lattice model is finite. This allows for the competition between the kinetic and potential terms of the Hamiltonian to drive a metal-insulator transition in the limit of $w\to\infty$ as well. Further, it has been argued in the DMFT literature that the Greens function matrix of the Hubbard model on the Bethe lattice with $w\to\infty$ becomes purely local (i.e., it has vanishing inter-site matrix elements). We can also see this from eqs.\eqref{local_gf},\eqref{nn_gf} where $(G^{H})^{-1}$ matrix has no $k$ dependence as $(G^{H})^{-1}_{ij}\propto 1/w \to 0$ as $w\to\infty$. Thus, in the self-energy 
\Sigma_{H} (\vec{k},\omega) &=& \omega - g_{0} + (t^{H} - g_{1})\xi_{\vec{k}}\nonumber\\
&\to & \lim_{w\to\infty} \omega - {g}_{0} + t^{H}\xi_{\vec{k}}~,
We can now see that even in the limit of $w\to\infty$, the competition between ${g}_{0}$ (the on-site repulsion) and the hopping related kinetic energy ($t^{H}\xi_{\vec{k}}$) can lead to a metal-insulator transition.
\par\noindent
While we have provided expressions for the single particle Greens functions of the $N$-site Hubbard model from their equivalent single particle Greens functions of the Hubbard dimer, we expect that eq.\eqref{green_eq_final} holds quite generally for the two-particle Greens function sector as well. We will provide these expressions at a later point in time. 
\section{Calculating $g_0$ and $g_1$ in practice}
The expressions for $g_0$ and $g_1$, as mentioned above, involve taking matrix elements of the Hubbard dimer Hamiltonian against the exact wavefunctions of the full Hubbard model.
	\item Since the full Hubbard model ground state wavefunctions are not readily available, one can, as the simplest approximation, assume the ground state has the form
		\ket{\tilde 0} = \sum_{\left<\tilde i \tilde j \right>}\ket{\Phi^D_{\tilde i\tilde j}}\otimes\ket{\Psi_{\hat{\tilde i}, \hat {\tilde j}}}\label{guess}
	$\ket{\Phi^D_{ij}}$ would be the ground state of the Hubbard dimer with $i,j$ creating the two sites of the dimer, while $\ket{\Psi_{\hat i, \hat j}}$ would be the wavefunction involving the rest of the sites. This ofcourse assumes that the two sets $(i,j)$ and $(a,b,...,h,k,...)$ are not entangled, and is not true in general. With this assumption, $g_0$ and $g_1$ become related to the inverse Greens functions of the Hubbard dimer. Such a choice of the wavefunction is motivated by the fact that in the auxiliary system we chose, the bath had only a diagonal interaction; there was no off-diagonal two particle scattering term. It further enforced when we extract the zero mode of the entire bath and keep just the zeroth site. The zeroth site, along with the impurity, then forms the $\ket{\Phi}$ part of the wavefunction.
	\item Another way of approaching this is to obtain the ground state wavefunction numerically, and then computing the matrix elements of $H^D$ against this wavefunction to obtain $g_0$ and $g_1$ for each value of $U^H$ and $t^H$ of the bare Hubbard model. This however means we lose any analytical insight into $g_0$ and $g_1$.
	\item The overall best way of calculating $g_0$ and $g_1$ is by systematically improving the ground state wavefunctions. This can done by obtaining the wavefunction of a Hamiltonian that has not just for the dimer, but also a bath (with dispersion) that connects with both the sites of the dimer:
		\label{dimer_dispersion}
	H^D_\text{bath}(\tilde i, \tilde j) = \underbrace{U\left(\tau_{\tilde i \uparrow}\tau_{\tilde i \downarrow} + \tau_{\tilde j \uparrow}\tau_{\tilde j \downarrow}\right) - t \sum_\sigma\left(c^\dagger_{\tilde i\sigma}c_{\tilde j\sigma} + \text{h.c.}\right)}_\text{Hubbard dimer} + \underbrace{\sum_{k\sigma}\epsilon_{k\sigma}\tau_{k\sigma}}_\text{bath} \underbrace{- t \sum_{k\sigma}\left(c^\dagger_{\tilde i\sigma}c_{k\sigma} + c^\dagger_{\tilde j\sigma}c_{k\sigma} + \text{h.c.}\right)}_\text{bath-dimer hybridisation}
	\includegraphics[width=0.5\textwidth]{./dimer_dispersion.png}
	\caption{Hubbard dimer with dispersion that connects to both sites}
By systematically increasing the number of nearest neighbours, we can get better and better ground state wavefunctions, with which we can compute matrix elements, and hence $g_0$, $g_1$. What this means is that if we had just a Hubbard dimer wavefunction, then the impurity can at most hop onto its nearest neighbour, and then hop back, hence the wavefunction in eq.~\ref{guess}. However, with this additional hopping, the impurity can now now hop on to the nearest neighbour $z$, and then on to the next nearest neighbours. This leads to entanglement between the impurity site and the next nearest neighbour sites, which was missing in eq.~\ref{guess}. The introduction of the dispersion gives the hope that we can capture a spectral function of the three peak form, because we have previously seen such a spectral function in the fixed point of the URG analysis of the SIAM (the fixed point Hamiltonian was an Anderson molecule with dispersion).
\item Another improvement can be made by introducing a self-energy into the dispersion. This introduces correlation on sites beyond the impurity and its nearest neighbour. The Hamiltonian would be the same as eq.~\ref{dimer_dispersion}, with $\epsilon_k$ replaced by $\tilde \epsilon_k \equiv \epsilon_k + \Sigma(k,\omega)$. The self-energy $\Sigma$ will have to be chosen depending on the phase we want to look at. With a $\Sigma$ that is singular near the Fermi surface at $\omega \to 0$, the dispersion will become gapped and the phase will be insulating. In such a situation, eq.~\ref{guess} will be a very good approximation to the actual ground state. On the other hand, if we use a self energy that vanishes near the Fermi surface ($\Sigma(\omega) \sim \omega^2$), then we will end up in a metallic phase. Introducing the self energy therefore gives us a larger variety of wavefunctions and features to work with.

@@
